A Survey of Scientific Workflow Management Systems

### 2014-04-21 paper.tex
more edits

    \item \textbf{Interoperability.} Interoperable workflows will benefit to a
        broad audience of users. Many efforts at this are made in the recent
        past, such as by the consurtium projects and by individual players.
        However, a truly interoperable set of standards for workflows are yet
        to emerge.

@@

 \item \textbf{catalog.} We study a number of important features as offered by
     contemporary SWEs. This will help catalog and collect the SWEs by their
     offerings and identify mechanisms and needs for new and existing features.
     This will also help identify current trends in scientific computing
     community.

 \item \textbf{application.} We identify the applications and application
     classes well adapted to the workflow systems and are benefitting by them.
     This will help readers relate to the application they are associated to.
     We hope that readers can make a choice of which SWE to use if there is an
     adaptation scope.

 \item \textbf{evolution.} We comment on the evolution of the existing workflow
     systems and the possible current and future development directions which
     will help readers be aware of the general development direction of the
     SWEs.

@@

%
\paragraph{Kepler/MoML} Kepler \cite{ludascher-altintas-etal:2005a} is a SWE
based upon the Ptollemy II framework \cite{eker-janneck-etal:2003} developed at
the University of California. Kepler workflows follow an actor-director
metaphor where an actor represents a unit of execution and the director
represents the paradigm for executing the workflows consisting of the actors.
Depending upon the types of directors, a Kepler workflow may exhibit a
different type of execution behavior. A workflow designer can chose from among
directors based upon the purpose of a workflow. Few of these include Process
Network (PN) for a parallel execution at the workflow level, Synchronous Data
Flow (SDF) for simple sequential data flow and Collection Oriented Modeling and
Design (COMAD) \cite{keplercomad} for a collection oriented workflow. Kepler
uses the Modeling Markup Language (MoML)~\cite{lee-neuendorffer:2000} for its
workflow representation. MoML is an extended XML syntax providing a rich
vocabulary to compose Kepler workflows. A workflow processor is called an
`actor' in Kepler.

Recent advances in Kepler include: MapReduce style computation, bioinformatics
computation, cloud computing~\cite{kepler2013}.

%
\paragraph{Dryad/--}

%
\paragraph{Pegasus/DAX} Pegasus \cite{deelman-peg} is a low-level distributed
computing framework that can map abstract workflows as DAGs to the concrete
underlying resources using intelligent and dynamic scheduling policies. A
Pegasus workflow is expressed in a simple XML-based language called DAX (DAG in
XML). DAX describes the logical workflow components, their inputs and outputs
and their links and dependencies in a Pegasus workflow. These kind of workflows
has been shown to enact optimally using Pegasus' smart memory and disk
management techniques. Pegasus manager buffers and maintains a fix-sized window
of executable tasks before accepting new tasks.

\paragraph{Wings/--} Wings is a Workflow system which does software choreography based on AI techniques~\cite{wings2013}
%
\paragraph{Swift/Swift Script} Swift \cite{swift} is a script-based scientific
workflow environment developed at the University of Chicago. A Swift workflow
specified in the SwiftScript scripting language is translated to an
intermediate XML-based representation and enacted by the Karajan workflow
enactment engine. The SwiftScript language is proposed as a workflow language
founded on a scripting approach programmers are familiar with. To make the
language adapted to grids, the inventors of SwiftScript generalized the use of
futures \cite{Halstead85multilisp:a} for every variables defined in the script.
Futures are non-blocking assignment variables, using a proxy to ensure
immediate execution of assignment and performing lazy blocking on variable
value accesses only. The systematic use of futures for every variables in the
language makes it completely data-driven: the execution progresses
asynchronously as long as a data access is not blocking. The availability of
data enables blocked thread to restart execution. The beauty of futures is to
make this process completely hidden from the programmer who can use a scripting
language he is familiar with without ever manipulating explicit parallel
programming constructs. Swift facilitates a highly asynchronous workflow
execution. Swift supports advanced workflow constructs such as parallel loops
and conditionals. Furthermore Swift supports processing of explicitly defined
arrays of data. Swift workflows take advantage of POSIX-style shared file
system to reuse the data produced by other processors.

%Recent advances in Swift: 
1. Swift has been shown to run on multiple clouds~\cite{ccgrid2013} via its
coasters mechanism~\cite{ucc2011}.

2. Swift/T~\cite{turbine2013}, runs scientific workflows over
HPC resources. Have been shown to work on petascale supercomputers such as the
IBM BlueGene and Cray series of machines.

3. GeMTC project investigating running many-task applications over GPUs via Swift/T.

@@

\paragraph{ASKALON/AGWL.} The ASKALON Problem Solving Environment (PSE)
\cite{fahringer-prodan-etal:2007} is developed at the University of Innsbruck.
ASKALON is described as a Scientific PSE capable of composing, hosting,
executing, sharing and reusing scientific workflows written in the Abstract
Grid Workflow Language (AGWL). AGWL is an XML-based workflow language that is
capable of expressing scientific application components chains at a high-level
of abstraction. A concrete workflow consists of the abstract workflow combined
with the resources and target execution environment binding the workflow. The
graphical interface of AGWL is based upon the UML (Unified Modeling Language)
diagrams.

\paragraph{The P-GRADE Portal/--} The P-GRADE Portal \cite{kacsuk-sipos:2005}
developed at the MTA SZTAKI Computation and Automation Institute, Budapest,
provides an access to grid computational and storage resources from a web
browser through a specialized workflow oriented grid portal. A workflow
development environment interfaces the users through this web based portal. The
portal interfaces the multiple grid computing middleware and storage resources
as a backend component of the environment. A highlight of P-GRADE portal is
that it provides access to multiple grids and scientific communities in a
single environment leading to a highly collaborative inter-community,
inter-grid research environment. An older version of P-GRADE used the
condor-Dagman as workflow engine while the recent version uses its own in-house
developed workflow engine called Xen.

%
\paragraph{DaGue/--}

@@

\paragraph{Galaxy Portal/JSON} Galaxy~\cite{blankenberg-vonkuster-etal:2010} is a
bioinformatics domain SWE based framework developed at the Penn State
University, Pennsilvania. It provides a webportal based environment to build
complex pipelines from an available repository of tools
(\textit{http://g2.bx.psu.edu}). Users can extend the tool repository provided
by Galaxy using their own Galaxy server or a web-server exposing the webservice
descriptions through public WSDLs. One of the major use of Galaxy is to achieve
a mix and match of data analysis tools in order to analyze a vast repository of
genome based database.

% == ENACTOR SCALABILITY AND OPTIMIZATION ==
\section{Enactor Scalability and Optimization} \label{sec:scal}
%We have results comparing \MOTEUR and Taverna performance!
%For Pegasus: Scaling up workflow based applications.
A simple definition of scalability in the context of SWE is that the
performance of the workflow engine increases linearly with the increasing
number of similar tasks.  Here, similar is an important term that means the
nature of tasks should be the same in terms of processing and data handling.
Depending upon the lag or increase in the linearity we say a workflow enactor
scales up or down.  In the present section, we focus on the scalability aspects
of the core workflow engine. We do not consider the specific scalability issues
related to a distributed infrastructure such as network latencies and protocol
offsets. However, there have been efforts to thwart such issues at the workflow
level by combining workflow enactment techniques with those of distributed
computing.  One such example is the map/reduce framework \cite{mapreduce}.
Recently, the map/reduce framework, that aims at performing distributed
processing on large datasets, has gained a significant popularity in workflow
community. It is based on the master/worker metaphor of distributed processing.
It employs the commonly used \textit{map} and \textit{reduce} functions from
the functional programming languages. The map function chops a large chunk of
data into smaller pieces and distributes it to \textit{workers} which in turn
can also chop it to further smaller datasets forming a tree-like hierarchy. The
worker process the data and return the results back to their parents. The
reduce step collects the results from the workers and combines them to obtain
the final result. One of the main reason for the popularity of the map/reduce
framework for workflows is the recent attention to collections based datasets
and processing. There exists implementations of the map/reduce framework
written in different languages with enhancements and support services such as a
sophisticated query service. Hadoop (\textit{http://hadoop.apache.org/}) is one
such implementation which has already been interfaced by the Kepler
SWE~\cite{wang-crawl-etal:2009}.

Large-scale workflows with thousands of tasks often pose memory and disk
management difficulties leading to scalability issues.  However, the
performance of a workflow engine is dependent on several factors including the
type of tasks and approach for submitting tasks for execution.  There is not
much research done towards the enactor scalability baring a couple of studies
carried out by C. Stratan et. al. \cite{stratan-iosup-etal:2008} and Deelman
et. al. \cite{callaghan-deelman-etal:2010}.  One way to compare the scalability
of different workflow environment is to experiment with a similar set of
workflows on a given system. In addition, different SWEs interface to different
DCIs. This is a time consuming process.  In the current study we do not
directly compare the workflow engine scalability but we study the similarities
and differences in the techniques employed by workflow engines in order to
achieve scalability.

@@

In the context of workflow enactors, scalability can be further classified in
terms of the computational and data transfer scalability. Computational
scalability could be expressed as the rate of increase in number of tasks
executed with respect to the increase in the elapsed time.

@@

Most workflow enactors exploit the parallelism at three-levels. First, at the
task-level wherein the independent branches of a workflow graph are enacted
concurrently. Second, at the data-level, where a same processor is instantiated
multiple times based upon the repeated operations on different datasets. Third,
the low-level parallelism that is encoded within the application components
independent of the workflow. At this level, the workflow enactor may
need to make a matching based upon the application type. For instance, a
workflow enactor may schedule the GPU-based processors on the resources that
indicate as possessing such capabilities.

In this section we analyze the scalability measures and techniques employed by
contemporary SWEs.

Illustrated below are some of the techniques used to achieve scalability in
various SWEs.

@@

Taverna, Swift-Karajan and Pegasus all provide a means to limit the maximum
number of processor instances that can run concurrently through a builtin
parameter. In Taverna, this can be configured at a workflow level while in
Pegasus, this can be configured at the Condor \cite{thain-tannenbaum-etal:2005}
scheduler level. In Swift-Karajan system this parameter is regulated
automatically as part of a ``throttling parallelfor" construct. A similar
configurable parameter is provided by the Knime thread manager in order to
limit the size of the thread-pool employed to run a workflow.

@@

In Swift systems the structure of a workflow is constructed and expanded
dynamically as opposed to that of Pegasus' static planning. A static workflow
graph allows optimizations based on the global state of the workflow, whereas a
dynamic workflow structure allows adaptations to changing environments.
Depending upon the execution environment and access conditions to the available
resources, each approach can result in the overall optimization of a given
application.

>>>>

    \item \textbf{Interoperability.} Interoperable workflows will benefit to a
        broad audience of users. Many efforts at this are made in the recent
        past, such as by the consurtium projects and by individual players.
        However, a truly interoperable set of standards for workflows are yet
        to emerge.
\end{enumerate}

@@

 \item \textbf{Catalog.} We study a number of important features as offered by
     contemporary SWEs. This will help catalog and collect the SWEs by their
     offerings and identify mechanisms and needs for new and existing features.
     This will also help identify current trends in scientific computing
     community.

 \item \textbf{Application.} We identify the applications and application
     classes well adapted to the workflow systems and are benefitting by them.
     This will help readers relate to the application they are associated to.
     We hope that readers can make a choice of which SWE to use if there is an
     adaptation scope.

 \item \textbf{Evolution.} We comment on the evolution of the existing workflow
     systems and the possible current and future development directions which
     will help readers be aware of the general development direction of the
     SWEs.

@@

%
\paragraph{Kepler/MoML} Kepler \cite{ludascher-altintas-etal:2005a} is a SWE
based upon the Ptollemy II framework \cite{eker-janneck-etal:2003} developed at
the University of California. Kepler workflows follow an actor-director
metaphor where an actor represents a processor and the director
represents the paradigm for executing the workflows consisting of the actors.
Depending upon the types of directors, a Kepler workflow may exhibit a
different type of execution behavior. A workflow designer can chose from among
directors based upon the purpose of a workflow. Few of these include Process
Network (PN) for a parallel execution at the workflow level, Synchronous Data
Flow (SDF) for simple sequential data flow and Collection Oriented Modeling and
Design (COMAD) \cite{keplercomad} for a collection oriented workflow. Kepler
uses the Modeling Markup Language (MoML)~\cite{lee-neuendorffer:2000} for its
workflow representation. MoML is an extended XML syntax providing a rich
vocabulary to compose Kepler workflows. A workflow processor is called an
`actor' in Kepler.

Recent advances in Kepler include: MapReduce style computation, bioinformatics
computation, cloud computing~\cite{kepler2013}.
\km{Todo: describe Kepler MapReduce in more detail, add Kepler's cloud abilities description}

%
\paragraph{Pegasus/DAX} Pegasus \cite{deelman-peg} is a low-level distributed
computing framework that can map abstract workflows as DAGs to the concrete
underlying resources using intelligent\km{elaborate} and dynamic\km{elaborate} scheduling policies. A
Pegasus workflow is expressed in a simple XML-based language called DAX (DAG in
XML). DAX describes the logical workflow components, their inputs and outputs
and their links and dependencies in a Pegasus workflow. These kind of workflows
has been shown to enact optimally using Pegasus' smart memory and disk
management techniques. Pegasus manager buffers and maintains a fix-sized window
of executable tasks before accepting new tasks. Recent developments in Pegasus
enables it to run workflows on cloud
environments~\cite{peg-cloud}.\km{elaborate, does not seem to do GPU
computing}.

\paragraph{Wings/--} Wings is a Workflow system which does software choreography based on AI techniques~\cite{wings2013}. \km{elaborate}.

%
\paragraph{Swift/Script, Tcl} Swift \cite{swift} is a script-based scientific
workflow environment developed at the University of Chicago and Argonne
National Laboratory. A Swift workflow specified in the Swift scripting language
and is translated to an intermediate XML-based representation and enacted by
the Karajan workflow enactment engine. The SwiftScript language is proposed as
a workflow language founded on a scripting approach that programmers are
familiar with. To make the language data parallel, Swift script generalizes the
use of futures \cite{Halstead85multilisp:a} for every variables defined in the
script.  Futures are non-blocking assignment variables, using a proxy to ensure
immediate execution of assignment and performing lazy blocking on variable
value accesses only. The systematic use of futures for every variables in the
language makes it completely data-driven: the execution progresses
asynchronously as long as a data access is not blocking. The availability of
data enables blocked thread to restart execution. Swift facilitates a highly
asynchronous workflow execution. Swift supports advanced workflow constructs
such as parallel loops and conditionals. Furthermore Swift supports processing
of explicitly defined arrays of data. Swift workflows take advantage of
POSIX-style shared file system to reuse the data produced by other processors.
Recent advances in Swift:
%Recent advances in Swift: 
\begin{enumerate}
    \item Swift has been shown to run on multiple
        clouds~\cite{ccgrid2013} via its coasters mechanism~\cite{ucc2011}.

    \item Swift/T~\cite{turbine2013}, runs scientific workflows over HPC
        resources. Have been shown to work on petascale supercomputers such as
        the IBM BlueGene and Cray series of machines.

    \item GeMTC project investigating running many-task applications over GPUs via Swift/T.
\end{enumerate}

@@

\paragraph{ASKALON/AGWL.} The ASKALON Problem Solving Environment (PSE)
\cite{fahringer-prodan-etal:2007} is developed at the University of Innsbruck.
ASKALON is described as a Scientific PSE capable of composing, hosting,
executing, sharing and reusing scientific workflows written in the Abstract
Grid Workflow Language (AGWL). AGWL is an XML-based workflow language that is
capable of expressing scientific application components chains at a high-level
of abstraction. A concrete workflow consists of the abstract workflow combined
with the resources and target execution environment binding the workflow. The
graphical interface of AGWL is based upon the UML (Unified Modeling Language)
diagrams.\km{Todo: current developments and advances.}

\paragraph{The P-GRADE Portal/--} The P-GRADE Portal \cite{kacsuk-sipos:2005}
developed at the MTA SZTAKI Computation and Automation Institute, Budapest,
provides an access to grid computational and storage resources from a web
browser through a specialized workflow oriented grid portal. A workflow
development environment interfaces the users through this web based portal. The
portal interfaces the multiple grid computing middleware and storage resources
as a backend component of the environment. A highlight of P-GRADE portal is
that it provides access to multiple grids and scientific communities in a
single environment leading to a highly collaborative inter-community,
inter-grid research environment. An older version of P-GRADE used the
condor-Dagman as workflow engine while the recent version uses its own in-house
developed workflow engine called Xen. \km{Investigate cloud and GPU capabilities}

@@

\paragraph{Galaxy Portal/JSON} Galaxy~\cite{blankenberg-vonkuster-etal:2010} is a
bioinformatics domain SWE based framework developed at the Penn State
University, Pennsilvania. It provides a webportal based environment to build
complex pipelines from an available repository of tools
(\textit{http://g2.bx.psu.edu}). Users can extend the tool repository provided
by Galaxy using their own Galaxy server. One of the major use of Galaxy is to
achieve a mix and match of data analysis tools in order to analyze a vast
repository of genome based database.

% == ENACTOR SCALABILITY AND OPTIMIZATION ==
\section{Enactor Scalability and Optimization} \label{sec:scal}
%We have results comparing \MOTEUR and Taverna performance!
%For Pegasus: Scaling up workflow based applications.
A simple definition of scalability in the context of SWE is that the
performance of the workflow engine increases linearly with the increasing
number of similar tasks. Depending upon the lag or increase in the linearity we
say a workflow enactor scales up or down.  In this section, we focus on
the scalability aspects of the workflow runtime. We do not consider the
specific scalability issues related to a distributed infrastructure such as
network latencies and protocol related overheads. However, there have been efforts to
thwart such issues at the workflow level by combining workflow enactment
techniques with those of distributed computing.  One such example is the
map/reduce framework \cite{mapreduce}.  Recently, the map/reduce framework,
that aims at performing distributed processing on large datasets, has gained a
significant popularity in workflow community. It is based on the master/worker
metaphor of distributed processing.  It employs the commonly used \textit{map}
and \textit{reduce} functions from the functional programming languages. The
map function chops a large chunk of data into smaller pieces and distributes it
to \textit{workers} which in turn can also chop it to further smaller datasets
forming a tree-like hierarchy. The worker process the data and return the
results back to their parents. The reduce step collects the results from the
workers and combines them to obtain the final result. One of the main reason
for the popularity of the map/reduce framework for workflows is the recent
attention to collections based datasets and processing. There exists
implementations of the map/reduce framework written in different languages with
enhancements and support services such as a sophisticated query service. Hadoop
(\textit{http://hadoop.apache.org/}) is one such implementation which has
already been interfaced by the Kepler SWE~\cite{wang-crawl-etal:2009}.

Large-scale applications with thousands of tasks often pose memory and disk
management difficulties leading to scalability issues.  However, the
performance of a workflow system is dependent on several factors including the
type of tasks and approach for submitting tasks for execution.  There is not
much research done towards the enactor scalability baring a couple of studies
carried out by C. Stratan et. al. \cite{stratan-iosup-etal:2008} and Deelman
et. al. \cite{callaghan-deelman-etal:2010}.  One way to compare the scalability
of different workflow environments is to experiment with a similar set of
workflows on a given system. In addition, different SWEs interface to different
DCIs. This is a time consuming process.  In the current study we do not
directly compare the workflow system scalability but we study the similarities
and differences in the techniques employed by workflow engines in order to
achieve scalability.

@@

In the context of workflow runtime, scalability can be further classified in
terms of data motion scalability. It can be expressed as the rate of increase
in number of data movement quantity with respect to the increase in the elapsed
time.

@@

Most workflow systems exploit the parallelism at three-levels. First, at the
task-level wherein the independent branches of a workflow graph are enacted
concurrently. Second, at the data-level, where a same processor is instantiated
multiple times based upon the repeated operations on different datasets. Third,
the low-level parallelism that is encoded within the application components
independent of the workflow. At this level, the workflow enactor may
need to make a matching based upon the application type. For instance, a
workflow enactor may schedule the GPU-based processors on the resources that
indicate as possessing such capabilities.

In this section we analyze the scalability measures and techniques employed by
contemporary SWEs. Illustrated below are some of the techniques used to achieve
scalability in various SWEs.

@@

Taverna, Swift-Karajan and Pegasus all provide a means to limit the maximum
number of processor instances that can run concurrently through a builtin
parameter. In Taverna, this can be configured at a workflow level while in
Pegasus, this can be configured at the Condor \cite{thain-tannenbaum-etal:2005}
scheduler level. In Swift-Karajan system this parameter is regulated
automatically as part of a ``throttling parallel foreach" construct. A similar
configurable parameter is provided by the Knime thread manager in order to
limit the size of the thread-pool employed to run a workflow.

@@

In some systems the structure of a workflow is constructed and expanded
dynamically as opposed to static planning. A static workflow graph allows
optimizations based on the global state of the workflow, whereas a dynamic
workflow structure allows adaptations to changing environments.  Depending upon
the execution environment and access conditions to the available resources,
each approach can result in the overall optimization of a given application.
